---
title: "An integrated population model for chum salmon in the lower Columbia River"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    toc_depth: 3
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tensorflow.one_based_extract = TRUE)
```

<br/>

[__Kale Bentley__](Kale.Bentley@dfw.wa.gov)  
_Fish Ecology and Life Cycle Monitoring Unit, Washington Department of Fish & Wildlife, Ridgefield, WA USA_

[__Thomas Buehrens__](Thomas.Buehrens@dfw.wa.gov)  
_Fish Ecology and Life Cycle Monitoring Unit, Washington Department of Fish & Wildlife, Ridgefield, WA USA_

[__Mark D. Scheuerell__](https://faculty.washington.edu/scheuerl/)  
_Northwest Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, WA USA_

***

__DISCLAIMER__  

This vignette is still in the testing and evaluating phase and should not be considered complete or error-free.

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

***

# Requirements

All analyses require the [R software](https://cran.r-project.org/) (v3.4+) for data retrieval, data processing, and summarizing model results, and the [Stan software](http://mc-stan.org/) (v2.17.1+) for Hamiltonian Monte Carlo (HMC) simulation.

We also need a few packages that are not included with the base installation of R, so we begin by installing them (if necessary) and then loading them.

```{r load_pkgs, message = FALSE, warning = FALSE}
## for file management
if(!require("here")) {
  install.packages("here")
  library("here")
}
## for reading
if(!require("readr")) {
  install.packages("readr")
  library("readr")
}
## for analysis
if(!require("rstan")) {
  install.packages("rstan")
  library("rstan")
}
## for model comparison
if(!require("loo")) {
  install.packages("loo")
  library("loo")
}
## set directory locations
datadir <- here("data")
# standir <- here("exec")
```

