---
title: 'Fitting Integrated Population Models to Lower Columbia River Chum Salmon Monitoring Data'
author: "Eric Buhle, Kale Bentley, Thomas Buehrens, Mark Scheuerell, Todd Hillson, and ..."
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: cerulean
    df_print: paged
    fig_caption: true
    toc: true
    toc_float: true
  word_document:
    toc: true
  pdf_document:
    highlight: haddock
    toc: yes
    number_sections: true
    toc_depth: '3'
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, highlight = TRUE, comment = NA, 
                      dev = "png", dev.args = list(type = "cairo-png"), dpi = 200,
                      out.width = "50%", fig.align = "center")

library(here) 
if(!require(captioner))
  devtools::install_github("adletaw/captioner")
library(captioner)
fig_nums <- captioner("Figure ", suffix = ": ", auto_space = FALSE, style = "b", style_prefix = TRUE)
library(kableExtra)
```
```{r width, include=FALSE}
options(width = 130)
```
```{r read_chunks, echo = FALSE}
knitr::read_chunk(here("analysis","R","02_LCRchumIPM_analysis.R"))
```

# Overview

Background on IPMs, outline of **salmonIPM**...

# Model Structure

## Process Model

### Egg Deposition and Egg-to-Smolt Survival

The chum life cycle begins with spawning, egg incubation and fry emergence, and shortly thereafter the downstream migration of juveniles (which we refer to as smolts). In **salmonIPM** we can fit three alternative spawner-recruit functions to describe the expected relationship between spawner abundance $S_{jt}$ and smolt abundance $M_{jt}$ in brood year $t$ in population $j$: density-independent discrete exponential, Beverton-Holt, and Ricker:

$$
f \left( S_{jt} | \alpha_{jt}, M_{\text{max},j}, A_{jt} \right) = 
\begin{cases}
\alpha_{jt} S_{jt} & \text{exponential} 
\\
\dfrac{ \alpha_{jt} S_{jt} }{ 1 + \alpha_{jt} S_{jt} / A_{jt} M_{\text{max},j} } 
& \text{Beverton-Holt} 
\\
\alpha_{jt} S_{jt} \text{exp}\left(- \dfrac{ \alpha_{jt}S_{jt} }{ \text{exp}(1) A_{jt} M_{\text{max},j} } \right) & \text{Ricker}
\end{cases}
$$

We use a nonstandard parameterization of the Ricker in terms of maximum smolt production or "capacity" $M_{\text{max}}$, corresponding to the mode of the function. This facilitates direct comparison with the Beverton-Holt (where $M_{\text{max}}$ is the asymptote) and is better-identified by data than the standard parameterization based on per capita density dependence. $M_{\text{max}}$ has units of density (fish per stream distance or area) and is then expanded to units of abundance based on habitat size $A_{jt}$. In the case of Lower Columbia chum, this is km of spawning habitat estimated using GIS.

Intrinsic productivity $\alpha_{jt}$ is calculated as a weighted mean of age-specific female fecundity $\mu_{E,a}$, weighted by the spawner age distribution $q_{jta}$, multiplied by the proportion of female spawners $q_{\text{F},jt}$ (discounted for the proportion of females that are not "green", i.e. not fully fecund, with discount rate $\delta_\text{NG} \in [0,1]$), and finally multiplied by the density-independent maximum egg-to-smolt survival $\psi_j$:

$$ 
\alpha_{jt} = \psi_j q_{F,jt} 
\left[p_{\text{G},jt} + \delta_\text{NG} \left(1 - p_{\text{G},jt} \right) \right] 
\sum_{a=3}^{5} q_{jta} \mu_{E,a}.
$$

The discount for partially spawned females is only relevant for one population, Duncan Channel, a constructed spawning channel in which some translocated females are believed to have already deposited eggs elsewhere. We assume the proportion of "green" females $p_{\text{G},jt}$ is known without error since translocated spawners are individually handled and visually assessed.

This formulation implicitly assumes egg deposition is density-independent while egg-to-smolt survival is density-dependent, resulting in realized survival $< \psi_j$ as spawner density increases. Maximum egg-to-smolt survival varies randomly among populations according to the hyperdistribution

$$
\text{logit}(\psi_j) \sim N(\mu_\psi, \sigma_\psi).
$$

Maximum smolt production varies randomly among populations according to the hyperdistribution

$$
\text{log}(M_{\text{max},j}) \sim 
N(\mu_{M_\text{max}}, \sigma_{M_\text{max}}).
$$

### Smolt Recruitment Process Error

Given the expected smolt production, realized smolt production (the unknown "true state") is lognormally distributed with process errors that include a common ESU-level autoregressive trend $\eta^\text{year}_{M,t}$ with first-order autocorrelation coefficient $\rho_{M}$ and innovation SD $\sigma^\text{year}_{M}$, plus unique independent shocks $\epsilon_{M,jt}$:

$$
\begin{aligned}
M_{jt} &= f \left( S_{jt} | \alpha_{jt}, M_{\text{max},j}, A_{jt} \right) \, 
\text{exp}( \eta^\text{year}_{M,t} + \epsilon_{M,jt} )
\\
\eta^\text{year}_{M,t} &\sim N(\rho_{M} \eta^\text{year}_{M,t-1}, \sigma^\text{year}_{M}) 
\\
\epsilon_{M,jt} &\sim N(0, \sigma_{M}).
\end{aligned}
$$

In **salmonIPM** we can include covariates of the parameters $\psi_j$ and $M_{\text{max},j}$ as well as the smolt recruitment process error term. We do not consider this option further here, but we anticipate incorporating environmental covariates in future development of the Lower Columbia chum IPM.

### Smolt-to-Adult Survival

Smolt-to-adult survival (SAR, $s_{MS}$) is assumed to be density-independent. The SAR process model for outmigrant cohort $t$ in population $j$ is logistic normal, including a common ESU-level first-order autoregressive trend $\eta^\text{year}_{MS,t}$ with first-order autocorrelation coefficient $\rho_{MS}$ and innovation SD $\sigma^\text{year}_{MS}$, plus unique independent shocks $\epsilon_{MS,jt}$:

$$
\begin{aligned}
\text{logit}( s_{MS,jt} ) &= \text{logit}( \mu_{MS} ) + 
\eta^\text{year}_{MS,t} + \epsilon_{MS,jt} 
\\
\eta^\text{year}_{MS,t} &\sim 
N(\rho_{MS} \eta^\text{year}_{MS,t-1}, \sigma^\text{year}_{MS}) 
\\
\epsilon_{MS,jt} &\sim N(0, \sigma_{MS}).
\end{aligned}
$$
As with the smolt recruitment parameters, **salmonIPM** can accommodate spatiotemporally varying covariates of $s_{MS}$, but we do not discuss this further here. 


### Conditional Age-at-Return

Adult age structure is modeled by defining a vector of conditional probabilities,
$\mathbf{p}_{jt} = [p_{3jt}, p_{4jt}, p_{5jt}] ^ \top$, where $p_{ajt}$ is the probability that an outmigrant in year $t$ in population $j$ returns at age $a$, given that it survives to adulthood. The unconditional probability is given by $s_{MS,jt} p_{ajt}$, where both SAR and $p_a$ are functions of underlying annual marine survival and maturation probabilities that are nonidentifiable without some ancillary data. This parameterization resolves the nonidentifiability.

The conditional age probabilities follow a logistic normal process model with hierarchical structure across populations and through time within each population. The additive log ratio,

$$
\text{alr}(\mathbf{p_{jt}}) = 
\left[ \text{log} \left( \dfrac{p_{3jt}}{p_{5jt}} \right), 
\text{log} \left( \dfrac{p_{4jt}}{p_{5jt}} \right) \right] ^ \top
$$

has a hierarchical bivariate normal distribution with population-level random effects $\boldsymbol{\eta}^\text{pop}_{\mathbf{p}, j}$ and unique residuals $\boldsymbol{\epsilon}_{\mathbf{p}, jt}$:

$$
\begin{aligned}
\text{alr}(\mathbf{p_{jt}}) &= \text{alr}(\boldsymbol{\mu}_\mathbf{p}) + 
\boldsymbol{\eta}^\text{pop}_{\mathbf{p}, j} +
\boldsymbol{\epsilon}_{\mathbf{p}, jt} 
\\
\boldsymbol{\eta}^\text{pop}_{\mathbf{p}, j} &\sim 
N(\mathbf{0}, \boldsymbol{\Sigma}^\text{pop}_\mathbf{p}) 
\\
\boldsymbol{\epsilon}_{\mathbf{p}, jt} &\sim
N(\mathbf{0}, \boldsymbol{\Sigma}_\mathbf{p}).
\end{aligned}
$$

Here the 2 $\times$ 2 covariances matrices $\boldsymbol{\Sigma}^\text{pop}_\mathbf{p}$ and $\boldsymbol{\Sigma}_\mathbf{p}$ allow correlated variation among age classes (on the unconstrained scale, not merely due to the simplex constraint on $\mathbf{p}$) across populations and through time within a population, respectively. For example, some populations or cohorts may skew overall younger or older than average. We parameterize each covariance matrix by a vector of standard deviations and a correlation matrix: 

$$
\begin{aligned}
\boldsymbol{\Sigma}^\text{pop}_\mathbf{p} &= 
\boldsymbol{\sigma}^\text{pop}_\mathbf{p}
\mathbf{R}_\mathbf{p}^\text{pop} 
{ \boldsymbol{\sigma}^\text{pop}_\mathbf{p} } ^ \top
\\
\boldsymbol{\Sigma}_\mathbf{p} &= 
\boldsymbol{\sigma}_\mathbf{p} 
\mathbf{R}_\mathbf{p}
\boldsymbol{\sigma}_\mathbf{p} ^ \top .
\end{aligned}
$$

### Sex Structure

Adult sex structure is modeled as the conditional probability $p_{\text{F},jt}$ that an outmigrant from population $j$ in year $t$ is female, given that it survives to adulthood. The proportion of females follows a process model that includes normally distributed population-specific random effects $\eta^\text{pop}_{\text{F}}$ with hyper-SD $\sigma^\text{pop}_\text{F}$ and unique residuals $\epsilon^\text{pop}_{\text{F}}$ with SD $\sigma_\text{F}$ around the hyper-mean $\mu_\text{F}$.

$$
\begin{aligned}
\text{logit}( p_{\text{F},jt} ) &= \text{logit}( \mu_\text{F} ) + 
\eta^\text{pop}_{\text{F},j} + \epsilon_{\text{F},jt} 
\\
\eta^\text{pop}_{\text{F},j} &\sim 
N(0, \sigma^\text{pop}_\text{F}) 
\\
\epsilon_{\text{F},jt} &\sim N(0, \sigma_\text{F})
\end{aligned}
$$

### Adult Recruitment and Composition

Recruitment of wild-origin adult spawners $S_\text{W}$ is calculated by summing over the corresponding outmigrant cohorts and subtracting the number of spawners removed for hatchery broodstock or to spawning channels ($B_{jt}$, assumed known without error). Fishery mortality of Lower Columbia chum is thought to be minimal.

$$
S_{\text{W}, jt} = \left(\sum_{a=3}^{5} 
s_{MS,j,t-a} \hspace{0.1cm} p_{aj,t-a} \hspace{0.1cm} M_{j,t-a} \right) - B_{jt} =
\left(\sum_{a=3}^{5} \tilde{S}_{\text{W}, ajt} \right) - B_{jt}.
$$

The proportion of naturally spawning hatchery-origin adults is modeled as an independent parameter in each population and year when hatchery programs were operational in the basin or when observed origin-frequency data indicate hatchery spawners were present. That is, we do not include a process model linking $p_{\text{HOS},jt}$ to hatchery releases and returns, although this is planned as a component of future model development. Abundance of hatchery-origin spawners is then calculated as

$$
S_{\text{H},jt} = \dfrac{ S_{\text{W},jt} \hspace{0.1cm} p_{\text{HOS},jt} } 
{ (1 - p_{\text{HOS},jt}) }.
$$

Total spawner abundance is then 

$$
S_{jt} = S_{\text{W},jt} + S_{\text{H},jt} + S^\text{add}_{jt}
$$ 
where $S^\text{add}_{jt}$ is the number of spawners that returned to populations $i \neq j$ and were translocated into population $j$. This is only nonzero in the case of Duncan Channel.

Spawner age structure is given by $\mathbf{q}_{jt} = [q_{3jt}, q_{4jt}, q_{5jt}]$, where $q_{ajt} = \tilde{S}_{\text{W},ajt} / \sum_a{\tilde{S}_{\text{W},ajt}}$. Spawner sex structure is given by the age-weighted average of female proportions from the respective outmigrant cohorts: $q_{\text{F},jt} = \sum_{a} {q_{ajt} \hspace{0.1cm} p_{\text{F},j,t-a}}$.


## Observation Model

### Fecundity

We modeled observations of fecundity from individual female chum salmon collected at hatcheries. The observation likelihood for the fecundity of female $i$ of age $a$ is a zero-truncated normal with age-specific mean and SD:

$$ E_{a,i}^\text{obs} \sim N_+(\mu_{E,a}, \sigma_{E,a}). $$

### Smolt and Spawner Abundance

Empirical estimates of smolt and spawner abundance come from monitoring programs that use various sampling methods (e.g., rotary screw traps for outmigrants and weirs or mark-recapture for adults) and statistical models (e.g., the **BTSPAS** R package or other Bayesian time-series models) to produce a posterior distribution of the abundance of a given life-history stage in each population and year. These methods have previously been applied and their output summarized by the posterior mean and SD. We used the posterior moments of abundance to compute the log-mean $\mu_{K,jt}$ and log-SD $\tau_{K,jt}$ for each life stage $K$ assuming the posterior is lognormal, which appears to be reasonable in general. We then used the observation-model posteriors as informative priors on the corresponding state variables in the IPM.

One complication is that outmigrant monitoring sites do not always have a one-to-one correspondence with independent populations. Specifically, the smolt trap in the mainstem Grays River intercepts juveniles produced in the mainstem population itself as well as the upstream West Fork Grays River and Crazy Johnson Creek populations, the latter of which is also sampled with a dedicated trap. To make the states comparable to the observations, we defined a derived state $\tilde{M}_{jt}$ that is the sum of smolts produced in population $j$ and all upstream populations.

$$
\begin{aligned}
\text{log}(\tilde{M}_{jt}) &\sim N(\mu_{\tilde{M},jt}, \tau_{\tilde{M},jt})
\\
\text{log}(S_{jt}) &\sim N(\mu_{S,jt}, \tau_{S,jt}).
\end{aligned}
$$

The IPM framework naturally handles missing observations, but in some cases an estimate of the observation prior log-mean $\mu_{K,jt}$ was available while the observation error log-SD $\tau_{K,jt}$ was missing or unknown. These missing values were imputed within the IPM by fitting a lognormal hyperdistribution to the known log-SDs:

$$
\begin{aligned}
\text{log}(\tau_{M,ij}) &\sim N( \mu_{\tau_M},  \sigma_{\tau_M})
\\
\text{log}(\tau_{S,ij}) &\sim N( \mu_{\tau_S},  \sigma_{\tau_S}).
\end{aligned}
$$

### Spawner Age Composition

Spawner age-composition data are collected from scale samples in each population and year (the IPM framework naturally handles cases in which no age samples are available). Age frequencies of wild spawners $\mathbf{n}_{ajt}^\text{obs} = [n_{3jt}^\text{obs}, n_{4jt}^\text{obs}, n_{5jt}^\text{obs}] ^\top$ are assumed to follow a multinomial observation likelihood with expected proportions given by the true state:

$$
\mathbf{n}_{ajt}^\text{obs} \sim 
\text{Multinomial} \left(  \sum_{a=3}^5 n_{ajt}^\text{obs},
\mathbf{q}_{jt} \right).
$$

### Spawner Rearing Type

Frequencies of wild- and hatchery-origin spawners are estimated based on otolith marking or genetic parentage-based tagging (PBT), and are assumed to follow a binomial observation likelihood given the total sample size assigned to rearing origin:

$$
n_{\text{H},jt}^\text{obs} \sim 
\text{Bin} \left( n_{\text{W},jt}^\text{obs} + n_{\text{H},jt}^\text{obs},
p_{\text{HOS},jt} \right).
$$

### Spawner Sex Composition

The observed frequency of female spawners is modeled with a binomial observation likelihood given the total sample size assigned to sex:

$$
n_{\text{F},jt}^\text{obs} \sim 
\text{Bin} \left( n_{\text{M},jt}^\text{obs} + n_{\text{F},jt}^\text{obs},
q_{\text{F},jt} \right).
$$


## Priors

Table of hyperpriors

$$
\begin{aligned}
\mu_{E,a} &\sim N_+(2500,500) \\
\sigma_{E,a} &\sim N_+(500,1000) \\
\delta_{\text{NG}} &\sim U(0,1) \\
\mu_\psi &\sim U(0,1) \\
\sigma_\psi &\sim N_+(0,1) \\
\mu_{M_\text{max}} &\sim N(m_{M_\text{max}}, s_{M_\text{max}}) \\
\sigma_{M_\text{max}} &\sim N_+(0,3) \\
\rho_M &\sim \text{PowerExp}_+(0,0.85,20) \\
\sigma^\text{year}_M &\sim N_+(0,2) \\
\sigma_M &\sim N_+(0,2) \\
\mu_{MS} &\sim U(0,1) \\
\rho_{MS} &\sim \text{PowerExp}_+(0,0.85,20) \\
\sigma^\text{year}_{MS} &\sim N_+(0,2) \\
\sigma_{MS} &\sim N_+(0,2) \\
\boldsymbol{\sigma}^\text{pop}_\mathbf{p} &\sim N_+(0,2) \\
\boldsymbol{\sigma}_\mathbf{p} &\sim N_+(0,2) \\
\mathbf{L}_\mathbf{p}^\text{pop} &\sim \text{LKJ}(1) \\
\mathbf{L}_\mathbf{p} &\sim \text{LKJ}(1) \\
\mu_\text{F} &\sim U(0,1) \\
\sigma^\text{pop}_\text{F} &\sim N_+(0,2) \\
\sigma_\text{F} &\sim N_+(0,2) \\
\mu_{\tau_M} &\sim N(0,1) \\
\sigma_{\tau_M} &\sim N_+(0,5) \\
\mu_{\tau_S} &\sim N(0,1) \\
\sigma_{\tau_S} &\sim N_+(0,5)
\end{aligned}
$$


```{r prior_table, results="asis", echo=FALSE, include=FALSE}
## alternative presentation
prior_table <- read.csv(here("analysis","prior_table.csv"))
knitr::kable(prior_table, escape = FALSE)
```


# Setup and Data

Load the packages we'll need...

```{r getting_started, message = FALSE, warning = FALSE}
```

Read in and manipulate the data...

```{r data, warning = FALSE, message = FALSE}
```

Let's look at the first few rows of `fish_data` to see the format **salmonIPM** expects...

```{r print_fish_data_SMS}
head(fish_data_SMS)
```


# Retrospective Models

Fit two-stage spawner-smolt-spawner models and explore output...

We fit exponential, Beverton-Holt and Ricker models, but model comparison using LOO is not feasible, so here we focus on the Ricker.

```{r fit_LCRchum_Ricker, eval = !exists("LCRchum_Ricker")}
```
```{r print_LCRchum_Ricker}
```


Plot estimated spawner-smolt production curves and parameters for the Beverton-Holt model.

```{r multiplot_LCRchum_Ricker, echo=FALSE, warning=FALSE, fig.width=12, fig.height=5.5, out.width="100%"}
mod_name <- "LCRchum_Ricker"
<<multiplot>>
```

`r fig_nums("plot_SR_LCRchum_Ricker", "Estimated Ricker spawner-recruit relationship (A, B) and intrinsic productivity (C) and capacity (D) parameters for the multi-population IPM. Thin lines correspond to each of 12 populations of Lower Columbia chum salmon; thick lines represent hyper-means across populations. In (A, B), each curve is a posterior median and the shaded region represents the 90% credible interval of the hyper-mean curve (uncertainty around the population-specific curves is omitted for clarity).")`

Here are the fits to the spawner data:

```{r plot_spawners_LCRchum_Ricker, echo = FALSE, warning=FALSE, fig.width=11, fig.height=7, out.width="100%"}
life_stage <- "S"
<<plot_spawner_smolt_ts>>
show(gg)
```

`r fig_nums("plot_spawners_LCRchum_Ricker", "Observed (points) and estimated spawner abundance for Lower Columbia River chum salmon populations. Filled points indicate known observation error SD, while SD for open points is imputed. The posterior median (solid gray line) is from the multi-population IPM. Posterior 90% credible intervals indicate process (dark shading) and observation (light shading) uncertainty.")`

And here are the fits to the much sparser smolt data:

```{r plot_smolts_LCRchum_Ricker, echo = FALSE, warning=FALSE, fig.width=11, fig.height=7, out.width="100%"}
life_stage <- "M"
<<plot_spawner_smolt_ts>>
show(gg)
```

`r fig_nums("plot_smolts_LCRchum_Ricker", "Observed (points) and estimated smolt abundance for Lower Columbia River chum salmon populations. Filled points indicate known observation error SD, while SD for open points is imputed. The posterior median (solid gray line) is from the multi-population IPM. Posterior 90% credible intervals indicate process (dark shading) and observation (light shading) uncertainty.")`

To understand how the IPM is imputing the observation error SD in cases where it is not reported, let's look at the lognormal hyperdistribution fitted to the known SD values...

```{r plot_obs_error_fit_LCRchum_Ricker, echo = FALSE, fig.width = 6, fig.height = 8}
<<plot_obs_error_fit>>
```

`r fig_nums("plot_obs_error_fit_LCRchum_Ricker", "Lognormal hyperdistributions used to impute unknown smolt and spawner observation error SDs in the IPM. The posterior median (line) and 90% credible interval (shading) of the distribution fitted to the known SD values (histogram) are shown for each life stage.")`

We can also compare the estimated spawner age-frequencies to the sample proportions from the BioData. Age composition varies quite a bit across populations and through time, reflecting fluctuations in cohort strength.

```{r plot_sex_ratio_ts_LCRchum_Ricker, echo = FALSE, warning=FALSE, fig.width=12, fig.height=7, out.width="100%"}
<<plot_sex_ratio_ts>>
show(gg)
```

`r fig_nums("plot_spawner_age_LCRchum_Ricker", "Observed (points) and estimated spawner age composition for Lower Columbia River chum salmon populations. The posterior distribution from the multi-population IPM is summarized by the median (solid line) and 90% credible interval (shading). The error bar around each observed proportion indicates the 90% binomial confidence interval based on sample size.")`

Sex ratio

```{r plot_spawner_age_LCRchum_Ricker, echo = FALSE, warning=FALSE, fig.width=11, fig.height=7, out.width="100%"}
<<plot_sex_ratio_ts>>
show(gg)
```

Proportion of hatchery-origin spawners

```{r plot_p_HOS_ts_LCRchum_Ricker, echo = FALSE, warning=FALSE, fig.width=11, fig.height=7, out.width="100%"}
<<plot_p_HOS_ts>>
show(gg)
```


# Forecasting

It is straightforward to use the IPM to generate forecasts of population dynamics...

```{r plot_spawners_LCRchum_Ricker_fore, echo = FALSE, warning=FALSE, fig.width=11.7, fig.height=7.2, out.width="100%", eval=FALSE}
mod_name <- "LCRchum_Ricker_fore"
life_stage <- "S"
<<plot_spawner_smolt_ts>>
```

`r fig_nums("plot_spawners_LCRchum_Ricker_fore", "Observed (points) and estimated spawner abundance for Lower Columbia River chum salmon populations, including 5-year forecasts. Filled points indicate known observation error SD, while SD for open points is imputed. The posterior median (solid gray line) is from the multi-population IPM. Posterior 90% credible intervals indicate process (dark shading) and observation (light shading) uncertainty.")`


Of course we could also look at forecasts of smolts, or any other state variable. Here are the 2020 forecasts of wild spawners for each population...

```{r spawner_forecast_table, echo = FALSE, eval=FALSE}
mod_name <- "LCRchum_Ricker_fore"
life_stage <- "S"   # "S" = spawners, "M" = smolts
<<forecast_df>>
kable(forecast_df, align = "l") %>% 
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```


