---
title: 'Fitting Integrated Population Models to Lower Columbia River Chum Salmon Monitoring Data'
author: "Eric Buhle, Kale Bentley, Thomas Buehrens, Mark Scheuerell, Todd Hillson, and ..."
date: "June 25, 2020"
output:
  html_document:
    keep_md: true
    df_print: paged
    fig_caption: true
    toc: true
    toc_float: true
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, highlight = TRUE, comment = NA, 
                      dev = "png", dev.args = list(type = "cairo-png"), dpi = 200,
                      out.width = "50%", fig.align = "center")

library(here)
if(!require(captioner))
  devtools::install_github("adletaw/captioner")
library(captioner)
fig_nums <- captioner("Figure", suffix = ": ", auto_space = FALSE, style = "b", style_prefix = TRUE)
```
```{r width, include=FALSE}
options(width = 130)
```
```{r read_chunks, echo = FALSE}
knitr::read_chunk(here("analysis","R","LCRchumIPM_analysis.R"))
```

# Overview

Background on IPMs, outline of **salmonIPM**...

[eqns]



# Setup and data

Load the packages we'll need...

```{r getting_started, message = FALSE, warning = FALSE}
```

Read in and manipulate the data...

```{r data}
```


# Retrospective models

Fit two-stage spawner-smolt-spawner models and explore output...

Density-independent

```{r fit_SMS_exp, eval = FALSE}
```
```{r print_SMS_exp}
```

Beverton-Holt

```{r fit_SMS_BH, eval = FALSE}
```
```{r print_SMS_BH}
```

Ricker

```{r fit_SMS_Ricker, eval = FALSE}
```
```{r print_SMS_Ricker}
```

Model comparison based on LOO (unhelpful because Pareto ks are too high)

```{r loo_SMS, warning = FALSE}
```

Plot estimated spawner-smolt production curves and parameters for the Beverton-Holt model.

```{r plot_SR_SMS_BH, echo=FALSE, fig.height=7, fig.width=7, warning=FALSE}
mod_name <- "SMS_BH"
<<plot_SR_params>>
```

Now do the same for the Ricker model.

```{r plot_SR_SMS_Ricker, echo = FALSE, warning = FALSE, fig.height = 7, fig.width = 7}
mod_name <- "SMS_Ricker"
<<plot_SR_params>>
```

The Ricker model is more biologically plausible, so let's proceed with that model for now. Here are the fits to the spawner data:

```{r plot_spawners_SMS_Ricker, echo = FALSE, fig.width=10, fig.height=6, out.width="100%"}
mod_name <- "SMS_Ricker"
life_stage <- "S"
<<plot_spawner_smolt_ts>>
```

And here are the fits to the much sparser smolt data:

```{r plot_smolts_SMS_Ricker, echo = FALSE, fig.width=10, fig.height=6, out.width="100%"}
mod_name <- "SMS_Ricker"
life_stage <- "M"
<<plot_spawner_smolt_ts>>
```

We can examine how the model partitions shared interannual fluctuations between the two life-stage transitions...

```{r plot_phi_SMS_Ricker, echo = FALSE, fig.width = 6, fig.height = 8}
mod_name <- "SMS_Ricker"
<<plot_phi>>
```


# Forecasts

It is straightforward to use the IPM to generate forecasts of population dynamics...

```{r plot_spawners_SMS_Ricker_fore, echo = FALSE, fig.width=10, fig.height=6, out.width="100%"}
mod_name <- "SMS_Ricker_fore"
life_stage <- "S"
<<plot_spawner_smolt_ts>>
```

